The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2013-08-21 sections/climate.tex
Irina's corrections in climate section

We force PISM with temperature and precipitation data from one observational dataset, three global and one regional reanalyses. Here we present this climate data and describe how we use it as a forcing.

\subsection{Observation data}

WorldClim is a high-resolution climate dataset built from observation data \citep{data:worldclim}. These climate surfaces were generated by spatial interpolation between a large set of weather stations over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data. This provides a resolution much higher than attained by general circulation models.

Figure~\ref{fig:temp} shows the spatial distribution of summer (JJA) and winter (DJF) air surface temperatures in WorldClim. Temperatures generally decrease from south to north and regions further inland experience colder winters. It should be noted that the temperature gradient is much stronger in the winter than it is in the summer. Temperature reach well above zero during the summer months over the entire modelling domain, except of the highest peaks. In other words, there is a strong contrast in seasonality between coastal and inland regions, even and regions where mean annual temperatures are well below freezing do experience warm summers.

Figure~\ref{fig:prec} shows the spatial distribution of DJF and JJA precipitation rates in WorldClim. Coastal regions generally receive much more precipitation than inland ones as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades. From a glacier mass-balance point of view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year. Although coastal regions experience most precipitation as snow during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.

@@

Throughout the simulation, the numerical model dynamically applies temperature lapse-rate corrections that account for differences between the climate forcing reference topography and the ice flow model topography on the one hand, and the evolution of ice thickness on the over hand. For reanalysis data, the circulation model surface topography (or surface geopotential height) was used as reference topography. For the WorldClim data, the aggregated hole-filled SRTM altitude field is used as reference topography. As a basal condition for the ice flow model, we use the ETOPO1\citep{data:etopo1} combined topography and bathymetry dataset. The reference topographies from the five forcing datasets, alongside the ETOPO1 data are shown in figure~\ref{fig:topo}. A lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations. No lapse-rate corrections apply to precipitation rates.

>>>>

We force PISM with temperature and precipitation data from one observational dataset, three global and one regional reanalyses. Here we present the climate datasets and describe how we use them as a forcing.

\subsection{Observational data}

WorldClim is a high-resolution climate dataset built from observational data \citep{data:worldclim}. These climate surfaces were generated by spatial interpolation between a large set of measurements taken at weather stations over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data. This provides a resolution much higher than attained by general circulation models in areas where sufficient amount of weather stations exist.

Figure~\ref{fig:temp} shows the spatial distribution of summer (JJA) and winter (DJF) air surface temperatures in WorldClim. In the study area, temperatures generally decrease from south to north and regions further inland experience colder winters. It should be noted that the temperature gradient is much stronger in the winter than it is in the summer. Temperature rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks. In other words, there is a strong contrast in seasonality between coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.

Figure~\ref{fig:prec} shows the spatial distribution of DJF and JJA precipitation rates in WorldClim. Coastal regions generally receive much more precipitation than inland ones as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades. From a glacier mass-balance point of view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year. Whereas coastal regions experience most precipitation as snow during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.

@@

Throughout the simulation, the numerical model dynamically applies temperature lapse-rate corrections that account for differences between the climate forcing reference topography and the ice flow model topography on the one hand, and the evolution of ice thickness on the over hand. For reanalysis data, the circulation model surface topography (or surface geopotential height) was used as reference topography. For the WorldClim data, the aggregated hole-filled SRTM altitude field is used as reference topography. As a basal condition for the ice flow model, we use the ETOPO1\citep{data:etopo1} combined topography and bathymetry dataset. The reference topographies from the five forcing datasets, alongside the ETOPO1 data are shown in figure~\ref{fig:topo}. An annual lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations. No lapse-rate corrections apply to precipitation rates.
